---
title: "Homework 7"
author: "your name here"
date: "Due Wed, Apr 6, 2022 at 11:00 PM in D2L"
urlcolor: blue
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
```


# Exercises

## 1.
Read the FlowingData article titled "How to Spot Visualization Lies" at [http://flowingdata.com/2017/02/09/how-to-spot-visualization-lies/](http://flowingdata.com/2017/02/09/how-to-spot-visualization-lies/). Summarize two key points in the article but **hint** pay attention to the entire article.


## 2.
Choose a data visualization from the [FlowingData](https://flowingdata.com/) or [Information is Beautiful](https://informationisbeautiful.net/) blogs on a topic of interest to you. Add a link to your chosen data visualization here:

[add your link here](add your link here)

Then answer the following questions:

a. Discuss one feature of the data visualization that adheres to good data visualization principles and explain why.
b. Discuss one feature of the data visualization that does not adhere to good data visualization principles or that could be improved and explain why.
c. What does the data visualization make you understand that, say, a written version of the same data likely wouldn’t?



## 3.
Select one of the data sets we have used in class (e.g., BaltimoreTowing.csv, Bike.csv, SeattleHousing.csv or one of the scraped data sets). Create a compelling graphic that adheres to the design principles from class. Include a written summary of the story your graphic displays and explain how it adheres to good data visualization design principles.