---
title: "Homework 2"
author: "your name here"
date: "Due Tuesday, February 7, 2022 at 5:00 PM in D2L"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn in both the PDF output and your R Markdown (.Rmd) file. Your .Rmd file should compile on its own if it is downloaded by your instructor.

## Load packages and data

_Reminder_: Packages that are not built-in to the original R installation need to first be installed in RStudio before the package will load into the current session. Refer to the homework instructions for package installation steps.

```{r load-packages, message = FALSE}
library(tidyverse)
library(dsbox)
```

## Exercises

### Exercise 1

Remove this text, and add your answer for Exercise 1 here.

### Exercise 2

Remove this text, and add your answer for Exercise 2 here.

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.

```{r prices-neighbourhoods}
# remove this comment and add the code for Exercise 3 here
```

### Exercise 4

Remove this text, and add your answer for Exercise 4 here.

```{r review-scores-rating}
# remove this comment and add the code for Exercise 4 here
```

### Exercise 5

Remove this text, and add your answer for Exercise 5 here.

```{r read-in-data}
# use the read_csv function to read in your data set here
```

```{r graphical-exploration}
# remove this comment and add the code for Exercise 5 here
```

