---
title: |
    | STAT 408 - Statistical Learning
    | Clustering
date: "December 7, 2017"
output: html_document
---


```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(randomForest)
library(maps)
library(plotrix)
library(mnormt)
library(rpart)
knitr::opts_chunk$set(echo = TRUE)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```


## Lab Exercise: Clustering Zoo Animals

Use the dataset create below for the following questions.

```{r}
zoo.data <- read.csv('http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/ZooClean.csv')
rownames(zoo.data) <- zoo.data[,1]
zoo.data <- zoo.data[,-1]
```

### a. MDS viz
Use multidimensional scaling to visualize the data in two dimensions.
```{r}
d <- dist(zoo.data) # euclidean distances between the rows
fit <- cmdscale(d, k=2) # k is the number of dim
fit.jitter <- fit + runif(nrow(fit*2),-.25,.25)
# plot solution 
plot(fit.jitter[,1], fit.jitter[,2], xlab="", ylab="",   main="",	type="n",axes=F)
box()
text(fit.jitter[,1], fit.jitter[,2], labels = rownames(zoo.data), cex=.8)
```

### b. MDS interpretation
What are two animals that are very similar and two that are very different?

### c. Hierarchical Clustering
Create a hierachical clustering object for this dataset.
```{r}
hc <- hclust(dist(zoo.data))
plot(hc, hang=-1)
```

Why are a leopard and raccoon clustered together for any cluster size?

### d. Cluster Viz
Now add colors corresponding to four different clusters to your MDS plot.

```{r}
plot(fit.jitter[,1], fit.jitter[,2], xlab="", ylab="",   main="",	type="n",axes=F)
box()
text(fit.jitter[,1], fit.jitter[,2], labels = rownames(zoo.data), col=cutree(hc,4), cex=.8)
```

Interpret what each of the four clusters correspond to.
