---
title: "Final Exam Review Key"
output: 
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, out.width = "75%")
library(tidyverse)
library(knitr)
```

## Coding and graphical questions

### Problem 1
Describe a way or sketch out R code to find the mean of the cost vector below. Note that `mean(cost)` will give an error.
```{r}
cost <- c("$1100", "$250.12", "$675")
```

General steps:

1. Remove `$` character
2. Convert to numeric
3. Take mean

Here is one possible solution:
```{r}
library(stringr)
cost <- str_remove(cost, "\\$")
cost
cost <- as.numeric(cost)
cost
mean(cost)
```


### Problem 2
Consider the following data frame:
```{r}
msu.football <- data.frame(opponent = c("Washington State", "South Dakota State", 
                                        "North Dakota", "Weber State", "Univ of Montana"),
                           points = c(0, 27, 49, 17, 134),
                           outcome = c("Loss", "Loss", "Win", "Loss", "Win"))
```
For each part below, explain what each line of code is doing (how each line of code helps produce the output). Then write the R output from the code below. Exactly one part will produce an error.

\newpage

a.
    ```{r}
for (i in 1:nrow(msu.football)){
  print(msu.football[i, 2])
}
```


b.
    ```{r}
msu.football %>% filter(outcome == "Loss") %>%
      summarize(MaxPoints = max(points))
```


c.
    ```{r, error = TRUE}
msu.football %>% select(points) %>%
  group_by(outcome) %>%
  count()
```


d.
    ```{r}
msu.football %>% ggplot(aes(x = points, y = outcome)) +
  geom_point() +
  labs(x = "Points Earned",
       y = "Outcome") +
  ggtitle("Points earned by outcome")
```

\newpage

e.
    ```{r}
msu.football %>% select(opponent) %>%
      mutate(CatGriz = case_when(
               opponent == "Univ of Montana" ~ "Yes",
               opponent != "Univ of Montana" ~ "No"
               )
             )
```


### Problem 3
Describe a strategy to merge the two data frames defined below without losing any information (i.e., keep all rows from `df1` and all rows from `df1`), then write the output you'd expect to see.
```{r}
df1 <- data.frame(school = c("MSU", "VT", "Mines", "Luther"),
                  state = c("MT", "VA", "CO", "IA"))
df2 <- data.frame(college = c("Mines", "MSU", "VT"),
                  enrollment = c(5794, 15688, 30598))
```

General steps:

1. Noticing that `school` and `college` serve as the variable on which to merge, rename one of the variables so that the two names match.
2. Perform a full join by `school`/`college`.


Here is one possible solution:
```{r}
df2 <- df2 %>% rename(school = college)
full_join(df1, df2)
```

\newpage

### Problem 4
You would like to write a function that will take our original
sample comprised of a single quantitative variable
and create a bootstrap confidence interval for the true population mean,
with a user-specified confidence level. Fill in the function below by adding R code or pseudocode wherever you see `<***>`.


One possible solution:
```{r}
bootstrap_means <- function(dat, conf.level = 0.95, reps = 1000) {
  # Function to generate a bootstrap distribution of means,
  # and calculate and report a confidence interval for the mean.
  # ARGS: 
  #   dat = sample data (numerical vector) 
  #   conf.level = confidence level as a decimal (defaults to 0.95)
  #   reps = number of bootstrap samples to generate (defaults to 1000)
  # RETURNS: a histogram plotting `reps` simulated means of size `sample_size`
  
  # Set up vector to store bootstrapped means
  means <- vector("numeric", length = reps)
  
  # Create bootstrap distribution of means
  for(i in 1:reps) {
    # Generate bootstrap sample
    ind <- sample(1:length(dat), length(dat), replace = TRUE)
    boot_samp <- dat[ind]
      
    # Calculate and store mean of bootstrap sample
    means[i] <- mean(boot_samp)
  }
  
  # Use distribution of bootstrapped means to calculate confidence interval
  quantile(means, c((1 - conf.level)/2, conf.level + (1 - conf.level)/2))
}

bootstrap_means(rnorm(50))
```

