---
title: 'STAT 408: Week 8'
subtitle: Strings, Dates and Factors
date: "3/8/2022"
output:
  ioslides_presentation:
    css: https://cdn.jsdelivr.net/npm/bootstrap@5.1.3/dist/css/bootstrap.min.css
    widescreen: yes
  beamer_presentation:
    theme: Berkeley
    colortheme: seahorse
    slide_level: 2
  revealjs::revealjs_presentation:
    transition: none
    incremental: no
  slidy_presentation:
    incremental: no
urlcolor: blue
---

<!-- To enable presenter mode add ?presentme=true to the URL of the presentation. For example: -->

<!-- my-presentation.html?presentme=true -->

<!-- The presenter mode window will open and will always re-open with the presentation until it is disabled with: -->

<!-- my-presentation.html?presentme=false -->

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, error=FALSE, message=FALSE)
```

# Baltimore Towing Data

## An Overview
This dataset contains information on vehicles towed in Baltimore, MD:

- A larger version of this dataset along with additional descriptions can be found at:  [https://data.baltimorecity.gov/Transportation/DOT-Towing/k78j-azhn](https://data.baltimorecity.gov/Transportation/DOT-Towing/k78j-azhn). 

<!-- Originally from data.gov -->

- The full version of the dataset contains 61,000 rows and 36 columns, where each row corresponds to a vehicle and the columns are information pertaining to the vehicle.

- We will be working with a smaller dataset with approximately 30,000 rows and 5 columns.

## The dataset
First read in the data set which is available at: [http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/BaltimoreTowing.csv](http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/BaltimoreTowing.csv).
```{r, message = F}
baltimore_tow <- read_csv('http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/BaltimoreTowing.csv')
str(baltimore_tow)
```

<!-- Vehicle type, make, model -->
<!-- What time vehicle placed into impound -->
<!-- Total paid to get vehicle out -->
<!-- Note in Make - all caps, misspelling -->
<!-- - some missing data -->
<!-- - form of date/time -->
<!-- - total paid in chr -->
<!-- - all variables characters -->

## Information for a few vehicles

```{r, echo = FALSE}
kable(head(baltimore_tow, 8)) %>%
   kable_styling()
```

# Data Wrangling Concepts

## Data Wrangling Concepts

- Dealing with strings
- Dealing with date/time objects
- Dealing with factors

# Goal 1. Average Towing Cost by Month

<!-- Think about how you would answer this with a data set like this -->

## Motivating Exercise: `group_by()`

Now compute the average towing cost grouped by month.

## Not Solution: `group_by()`

Now compute the average towing cost grouped by month.

```{r, eval = F}
baltimore_tow %>% 
  group_by(month) %>% 
  summarize(mean.cost = mean(totalPaid))
```

<!-- We don't have all the data in the format we need to solve the problem -->
<!-- We need month - part of date  -->
<!-- How would you do this if you didn't know how to write code (e.g., in Excel) -->


## Motivating Exercise: `group_by()`

Now compute the average towing cost grouped by month.


```{r, echo = FALSE}
kable(head(baltimore_tow, 4)) %>% kable_styling()
```


##  `substr()` function
Consider adding a column for year to the data set. This can be done using `substr().`

*Usage*: substr(x, start, stop)

*Arguments*: 

- x, text	a character vector.
- start, first integer. The first element to be extracted
- stop, last integer. The last element to be extracted

## Exercise: Using the `substr()` function
Use the `substr()` function to extract month and create a new variable in R. 

## Solution: Using the `substr()` function
Use the `substr()` function to extract month and create a new variable in R. 

```{r}
baltimore_tow$month <- substr(baltimore_tow$receivingDateTime, 0, 2)
head(baltimore_tow$month)
```

<!-- Note character vector -->

## Motivating Exercise: `group_by()`

Now compute the average towing cost grouped by month.

```{r, warning = F}
baltimore_tow %>% 
  group_by(month) %>% 
  summarize(mean.cost = mean(totalPaid))
```

<!-- Note NAs - why? -->
<!-- possible some NAs in the values given to the function -->
<!-- - in this case: dollar signs makes variable a character -->

## `strsplit()` function
In many situations, the year could be in a different position so the `substr()` might not work. For example month the date could be coded `4/1/2015` rather than `04/01/2015` So consider, using `strsplit()` instead.

*Usage*: strsplit(x, split)

*Arguments*:

- x: character vector, each element of which is to be split. Other inputs, including a factor, will give an error.
- split: character vector (or object which can be coerced to such) containing regular expression(s) (unless fixed = TRUE) to use for splitting. 

## Exercise: Using the `strsplit()` function
Use the `strsplit()` function to remove the dollar sign from the cost.

## Solution: Using the `strsplit()` function
Use the `strsplit()` function to remove the dollar sign from the cost.

```{r}
## example for one row
strsplit(baltimore_tow$totalPaid[1],'$', fixed = T)[[1]][2]
```

<!-- Run inner part of code first -->
<!-- Result is a list -->

# Lists

## Data structure overview (review)
The base data structures in R can be organized by dimensionality and whether they are homogenous.

Dimension     | Homogeneous     | Heterogeneous
------------- | -------------  | -------------
1d            | Vector         | List
2d            | Matrix         | Data Frame
no d          | Array          |

## Lists
Consider the two lists 

```{r}
msu.info <- list(
  name = c('Waded Cruzado', 'Stacey Hancock'), 
  degree.from = c('University of Texas at Arlington', 
                  'Colorado State University'),
  job.title = c('President',
                'Associate Professor of Statistics'))

msu.info2 <- list(
  c('Waded Cruzado',
    'University of Texas at Arlington',
    'President'), 
  c('Stacey Hancock',
    'Colorado State University',
    'Associate Professor of Statistics'))
```

## List Output

```{r}
msu.info

msu.info2
```

## Lists - indexing
With the current lists we can index elements using the double bracket `[[ ]]` notation or if names have been initialized, those can be used too.

So the first element of each list can be indexed
```{r, mysize=TRUE, size='\\tiny'}
msu.info[[1]]
msu.info$name
```

## Exercise: Lists 
Explore the indexing with these commands.
```{r, eval=F}
msu.info[1]
msu.info[[1]]
msu.info$name[2]
msu.info[1:2]
unlist(msu.info)
```

## Indexing lists

> "If list `x` is a train carrying objects, then `x[[5]]` is the object in car 5; `x[4:6]` is a train of cars 4-6."

— @RLangTip

Source: http://adv-r.had.co.nz/Subsetting.html

## Elements of lists need not be the same class or even dimension!

```{r}
list(c("Jan","Feb","Mar"), 
     matrix(c(3,9,5,1,-2,8), nrow = 2),
     list("green", 12.3)
     )
```

## Solution: Using the `strsplit()` function (revisited)
Use the `strsplit()` function to remove the dollar sign from the cost.

```{r}
strsplit(baltimore_tow$totalPaid[1:2], '$', fixed = T)[[1]][2]
```


# lubridate package

## lubridate package

lubridate is a tidyverse package for manipulating date objects. There is a nice [website](https://lubridate.tidyverse.org) with a cheatsheet.

<!-- website is first place to go for questions -->

## Date objects for Baltimore tow

```{r, message=F}
library(lubridate) # loads with tidyverse
class(baltimore_tow$receivingDateTime)
baltimore_tow <- baltimore_tow %>% 
  mutate(date_time = mdy_hms(receivingDateTime))
class(baltimore_tow$date_time)
```

<!-- Make character a date_time object -->

## Date objects for Baltimore tow

```{r}
head(month(baltimore_tow$date_time))
head(year(baltimore_tow$date_time))
```

<!-- Only real challenge is getting to be a date_time object in the first place -->

## Stringr Package
The `stringr` package ([cheat sheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf)) provides a nice set of tools. There is also an [information page](https://stringr.tidyverse.org/articles/from-base.html).


## Exercise: Stringr approach
Use the `stringr` package to remove (replace) the dollar sign. Note that a dollar sign is a special character, so you'll need to use `\\$`.

<!-- Open up website and cheatsheet and find a function we can use -->

## Solution: Stringr approach
Use the `stringr` package to remove (replace) the dollar sign
```{r}
library(stringr)
baltimore_tow$cost <- 
  as.numeric(str_replace(baltimore_tow$totalPaid, '\\$',''))
```

<!-- convert character string into a number -->

## Motivating Exercise: `group_by()`

Now compute the average towing cost grouped by month.

```{r}
baltimore_tow %>% 
  group_by(month) %>% 
  summarize(mean.cost = mean(cost), .groups = 'keep')
```

<!-- The challenge wasn't the dplyr code, but how to get "month" and "cost" in a way we can use them -->

# Goal 2. Type of Vehicles Towed by Month


## Goal 2. Type of Vehicles Towed by Month
Next we wish to compute how many vehicles were towed for each vehicle type.

However, we want to take a close look at the vehicle types in the data set and perhaps create more useful groups.

## `unique` function -- how to group vehicles
First examine the unique types of vehicles in this data set.

```{r}
unique(baltimore_tow$vehicleType)
```

<!-- dplyr function to do this? -->

## Grouping
First consider reasonable groups for vehicle types.

1. Cars - (Car, convertible)
2. Large Cars - (SUV, Station Wagon, Sport Utility Vehicle, Van, Taxi)
3. Trucks - (Pick-up Truck, Pickup Truck)
4. Large Trucks - (Truck, Tractor Trailer, Tow Truck, Tractor, Construction Equipment, Commercial Truck)
5. Bikes - (Motor Cycle (Street Bike), Dirt Bike, All terrain - 4 wheel bike, Mini-Bike)
6. Misc (delete) - (Boat, Golf Cart, Trailer)

## Messy Data: Grouping
Next examine values in some of these groups, we will just look at the vehicle type of 'Truck'.

```{r, mysize=TRUE, size='\\scriptsize', tidy=T}
unique(baltimore_tow$vehicleMake[baltimore_tow$vehicleType == 'Truck'])
```

<!-- Work your way from inner-most code to outer-most -->
<!-- How would you do this in tidyverse? -->

Note that there are several spelling errors in this data set. How do we combine them?


## Messy Data: Data Cleaning
Spelling errors can be addressed, by reassigning vehicles to the correct spelling.
```{r}
baltimore_tow$vehicleMake[baltimore_tow$vehicleMake == 
                            'Peterbelt'] <- 'Peterbilt'
baltimore_tow$vehicleMake[baltimore_tow$vehicleMake == 
                            'Internantional'] <- 'International'
baltimore_tow$vehicleMake <-
  str_replace(baltimore_tow$vehicleMake,'Izuzu','Isuzu')
baltimore_tow$vehicleMake <-
  str_replace(baltimore_tow$vehicleMake,'Frightliner','Freightliner')
```
Also note that many of the groupings have mis-classified vehicles, but we will not focus on that yet.

<!-- "Traditional" way to do it -->

## Exercise: Delete Misc. Type Vehicles

First we will delete golf carts, boats, and trailers. There are several ways to do this, consider making a new data frame called `balt_tow_small` that does not include golf carts, boats, and trailers.

## Solution: Delete Misc. Type Vehicles

First we will delete golf carts, boats, and trailers.

```{r}
balt_tow_small <- baltimore_tow %>% 
  filter(!(vehicleType %in% c("Golf Cart", "Boat", "Trailer")))
```


## Exercise: Create Additional Groups

Now we need to create a variable for the additional groups below.

1. Cars - (Car, convertible)
2. Large Cars - (SUV, Station Wagon, Sport Utility Vehicle, Van, Taxi)
3. Trucks - (Pick-up Truck, Pickup Truck)
4. Large Trucks - (Truck, Tractor Trailer, Tow Truck, Tractor, Construction Equipment, Commercial Truck)
5. Bikes - (Motor Cycle (Street Bike), Dirt Bike, All terrain - 4 wheel bike, Mini-Bike)

## Solution: Create Additional Groups

One way to create groups is by creating a new variable

```{r}
balt_tow_small$Group <- '' # Creates empty string for all rows in data set

balt_tow_small$Group[balt_tow_small$vehicleType %in% 
                       c('Car','Convertible')] <- 'Cars'

balt_tow_small$Group[balt_tow_small$vehicleType %in% c('SUV',
 'Station Wagon','Sport Utility Vehicle','Van','Taxi')] <- 'Large Cars'

balt_tow_small$Group[balt_tow_small$vehicleType %in% 
                       c('Pick-up Truck','Pickup Truck')] <- 'Trucks'

balt_tow_small$Group[balt_tow_small$vehicleType %in% 
    c('Truck','Tractor Trailer','Tow Truck','Tractor',
      'Construction Equipment','Commercial Truck')] <- 'Large Trucks'

balt_tow_small$Group[balt_tow_small$vehicleType %in% 
    c('Motor Cycle (Street Bike)','Dirt Bike','Mini-Bike',
      'All terrain - 4 wheel bike')] <- 'Bikes'
```


## Solution: 
Next we wish to compute how many vehicles were towed for each vehicle type

```{r}
balt_tow_small %>% count(Group)
```

# Factors

<!-- Easier way to do this - but we need to talk about factors -->

## Factors...
Factors are a specific way to store categorical data. Using factors results in a more efficient data storage process, but can be cumbersome.

Factors can be necessary for making plots and fitting models in R.



## forcats Package
The `forcats` package, [website](https://forcats.tidyverse.org), is a tidyverse package designed for dealing with categorical factors.

## Character Values

```{r}
favorite_day <- c('Friday', 'Saturday', 'Sunday', 'Tuesday', 'Saturday', 'Saturday')
class(favorite_day)
```

## Creating Factors

```{r}
day_factor <- as.factor(favorite_day)
class(day_factor)
sort(day_factor)
```

<!-- as.... -> coerce data to be a different class type -->
<!-- - you'd hope sort would sort in order of weekdays -->
<!-- - alphabetical -->

## Reordering Factors

```{r}
library(forcats)
day_factor <- fct_relevel(day_factor,
                          c('Sunday','Tuesday','Friday','Saturday'))
sort(day_factor)
```


## Creating Factors
Rather than coercing a class variable to be a factor, the factor can be created directly.
```{r}
day_factor2 <- factor(c('Friday', 'Saturday', 'Sunday', 'Monday'),
                      levels = c('Sunday', 'Monday', 'Tuesday', 
                        'Wednesday','Thursday','Friday','Saturday'))
sort(day_factor2)
```

## Collapsing Factors
Factors can also easily be collapsed with `forcats`

```{r}
balt_tow_small %>% 
  mutate(Group2 = fct_collapse(vehicleType, 
           Cars = c('Car','Convertible'),
           Large_Cars = c('SUV', 'Station Wagon',
                          'Sport Utility Vehicle',
                          'Van','Taxi'),
           Trucks = c('Pick-up Truck','Pickup Truck'),
           Large_Trucks = c('Truck', 'Tractor Trailer',
                            'Tow Truck', 'Tractor',
                            'Construction Equipment',
                            'Commercial Truck'),
           Bikes = c('Motor Cycle (Street Bike)', 'Dirt Bike',
                     'Mini-Bike','All terrain - 4 wheel bike')
           )
         ) %>% 
  mutate(Group2 = fct_infreq(Group2)) %>%
  group_by(Group2) %>% 
  summarize(ave_cost = mean(cost), .groups = 'drop')
```
