---
title: "Week 8 Lecture Exercises"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Baltimore Towing Data

## The dataset

```{r, message = F}
baltimore_tow <- read_csv('http://www.math.montana.edu/ahoegh/teaching/stat408/datasets/BaltimoreTowing.csv')
str(baltimore_tow)
```

## Information for a few vehicles

```{r}
kable(head(baltimore_tow, 8)) %>%
   kable_styling()
```

# Data Wrangling Concepts

# Goal 1. Average Towing Cost by Month

Now compute the average towing cost grouped by month.

## Not Solution: group_by()

```{r, eval = F}
baltimore_tow %>% 
  group_by(month) %>% 
  summarize(mean.cost = mean(totalPaid))
```


## Exercise: Using the `substr()` function
Use the `substr()` function to extract month and create a new variable in R. 

## Exercise: Using the `strsplit()` function
Use the `strsplit()` function to remove the dollar sign from the cost.

## Lists
Consider the two lists 

```{r}
msu.info <- list(
  name = c('Waded Cruzado', 'Stacey Hancock'), 
  degree.from = c('University of Texas at Arlington', 
                  'Colorado State University'),
  job.title = c('President',
                'Associate Professor of Statistics'))

msu.info2 <- list(
  c('Waded Cruzado',
    'University of Texas at Arlington',
    'President'), 
  c('Stacey Hancock',
    'Colorado State University',
    'Associate Professor of Statistics'))
```

## List Output

```{r}
msu.info

msu.info2
```

## Lists - indexing
```{r}
msu.info[[1]]
msu.info$name
```

## Exercise: Lists 
Explore the indexing with these commands.
```{r, eval=F}
msu.info[1]
msu.info[[1]]
msu.info$name[2]
msu.info[1:2]
unlist(msu.info)
```

## Elements of lists need not be the same class or even dimension!

```{r}
list(c("Jan","Feb","Mar"), 
     matrix(c(3,9,5,1,-2,8), nrow = 2),
     list("green", 12.3)
     )
```

# lubridate package

## Date objects for Baltimore tow

```{r, message=F}
library(lubridate) # loads with tidyverse
class(baltimore_tow$receivingDateTime)
baltimore_tow <- baltimore_tow %>% 
  mutate(date_time = mdy_hms(receivingDateTime))
class(baltimore_tow$date_time)
```

## Date objects for Baltimore tow

```{r}
head(month(baltimore_tow$date_time))
head(year(baltimore_tow$date_time))
```


## Exercise: Stringr approach
Use the `stringr` package to remove (replace) the dollar sign. Note that a dollar sign is a special character, so you'll need to use `\\$`.

# Goal 2. Type of Vehicles Towed by Month

## `unique` function -- how to group vehicles
First examine the unique types of vehicles in this data set.

```{r}
unique(baltimore_tow$vehicleType)
```

## Grouping
First consider reasonable groups for vehicle types.

1. Cars - (Car, convertible)
2. Large Cars - (SUV, Station Wagon, Sport Utility Vehicle, Van, Taxi)
3. Trucks - (Pick-up Truck, Pickup Truck)
4. Large Trucks - (Truck, Tractor Trailer, Tow Truck, Tractor, Construction Equipment, Commercial Truck)
5. Bikes - (Motor Cycle (Street Bike), Dirt Bike, All terrain - 4 wheel bike, Mini-Bike)
6. Misc (delete) - (Boat, Golf Cart, Trailer)

## Messy Data: Data Cleaning
Spelling errors can be addressed, by reassigning vehicles to the correct spelling.
```{r}
baltimore_tow$vehicleMake[baltimore_tow$vehicleMake == 
                            'Peterbelt'] <- 'Peterbilt'
baltimore_tow$vehicleMake[baltimore_tow$vehicleMake == 
                            'Internantional'] <- 'International'
baltimore_tow$vehicleMake <-
  str_replace(baltimore_tow$vehicleMake,'Izuzu','Isuzu')
baltimore_tow$vehicleMake <-
  str_replace(baltimore_tow$vehicleMake,'Frightliner','Freightliner')
```
Also note that many of the groupings have mis-classified vehicles, but we will not focus on that yet.

## Exercise: Delete Misc. Type Vehicles

First we will delete golf carts, boats, and trailers. There are several ways to do this, consider making a new data frame called `balt_tow_small` that does not include golf carts, boats, and trailers.

## Exercise: Create Additional Groups

Now we need to create a variable for the additional groups below.

1. Cars - (Car, convertible)
2. Large Cars - (SUV, Station Wagon, Sport Utility Vehicle, Van, Taxi)
3. Trucks - (Pick-up Truck, Pickup Truck)
4. Large Trucks - (Truck, Tractor Trailer, Tow Truck, Tractor, Construction Equipment, Commercial Truck)
5. Bikes - (Motor Cycle (Street Bike), Dirt Bike, All terrain - 4 wheel bike, Mini-Bike)


## Solution: 
Next we wish to compute how many vehicles were towed for each vehicle type

```{r}
balt_tow_small %>% count(Group)
```

# Factors
## Character Values

```{r}
favorite_day <- c('Friday', 'Saturday', 'Sunday', 'Tuesday', 'Saturday', 'Saturday')
class(favorite_day)
```

## Creating Factors

```{r}
day_factor <- as.factor(favorite_day)
class(day_factor)
sort(day_factor)
```

## Reordering Factors

```{r}
library(forcats)
day_factor <- fct_relevel(day_factor,
                          c('Sunday','Tuesday','Friday','Saturday'))
sort(day_factor)
```


## Creating Factors
Rather than coercing a class variable to be a factor, the factor can be created directly.
```{r}
day_factor2 <- factor(c('Friday', 'Saturday', 'Sunday', 'Monday'),
                      levels = c('Sunday', 'Monday', 'Tuesday', 
                        'Wednesday','Thursday','Friday','Saturday'))
sort(day_factor2)
```

## Collapsing Factors
Factors can also easily be collapsed with `forcats`

```{r}
balt_tow_small %>% 
  mutate(Group2 = fct_collapse(vehicleType, 
           Cars = c('Car','Convertible'),
           Large_Cars = c('SUV', 'Station Wagon',
                          'Sport Utility Vehicle',
                          'Van','Taxi'),
           Trucks = c('Pick-up Truck','Pickup Truck'),
           Large_Trucks = c('Truck', 'Tractor Trailer',
                            'Tow Truck', 'Tractor',
                            'Construction Equipment',
                            'Commercial Truck'),
           Bikes = c('Motor Cycle (Street Bike)', 'Dirt Bike',
                     'Mini-Bike','All terrain - 4 wheel bike')
           )
         ) %>% 
  mutate(Group2 = fct_infreq(Group2)) %>%
  group_by(Group2) %>% 
  summarize(ave_cost = mean(cost), .groups = 'drop')
```
