---
title: "Homework 9"
author: "your name here"
date: "Due Wed, May 4, 2022 at 11:00 PM in D2L"
urlcolor: blue
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Data

For this homework, you will use the `NHANES` data set in the `NHANES` package.
You will need to install the `NHANES` package, then type `?NHANES` to view
a description of the data set.

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse)
library(NHANES) # You may need to install this package
library(skimr)
```


# Exercises

## 1. Examine Data Set

Before any modeling occurs, it is a good idea to examine the data set to get
a sense of what the variables in the data set look like. We've seen the `summary`
and `str` functions in base R, but another nice function to view a quick summary
of a data frame is the `skim` function in the `skimr` library. 

Use the `skim` function to print a summary of the data set. Write a few sentences
commenting on a few interesting aspects of this data set.
(For example, only 88 observations have a value for `HeadCirc` -- the rest
are missing. Thus, this variable will be of little value when fitting models.)

## 2. Create Training / Testing Data

Run the code below to separate the `NHANES` data set at random into 
70% training and 30% testing.
The `set.seed` function ensures that everyone is using the same
training and testing data set.

```{r}
set.seed(05042022)
num.obs <- nrow(NHANES)
test.ids <- base::sample(1:num.obs, size=round(num.obs*.3))
test.set <- NHANES[test.ids,]
train.set <- NHANES[(1:num.obs)[!(1:num.obs) %in% test.ids],]
```


## 3. Supervised Learning

The `NHANES` data set contains a binary variable `SleepTrouble` that indicates whether each person has trouble sleeping. For each of the following models below,

- Build a classifier for `SleepTrouble` using your training data set
- Report its effectiveness on your testing data set
- Make an appropriate visualization of the model
- Interpret the results - what have you learned about people's sleeping habits?

a. Logistic regression
b. Decision tree
c. Random forest

You may use whatever variables you like, except for `SleepHrsNight` and `ID`.


## 4. Unsupervised Learning

a. Compare and contrast hierarchical clustering with k-means clustering.

b. Describe a situation where clustering could be useful.

## Extra Credit

The `nasaweather` package contains data about tropical storms from 1995–2005
in the data set `storms`.

a. Create a scatterplot of these storms with wind speed on the y-axis, 
pressure on the x-axis, and colored by type of storm.

b. Use a k-means clustering algorithm to cluster these data based on the
two variables `wind` and `pressure` with four clusters. Describe
and visualize how well your clusters match up to the types of storms
given in the data set.



# Cite Sources

Remove this text and cite the sources used on your homework here.
