---
title: "STAT 408: Week "
subtitle: "Teach This Sometime"
date: "someday"
urlcolor: blue
output: 
  ioslides_presentation:
    widescreen: true
  beamer_presentation:
    theme: "Berkeley"
    colortheme: "seahorse"
    slide_level: 2
  revealjs::revealjs_presentation:
    transition: none
    incremental: false
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, error=FALSE, message=FALSE)
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})
```

# Lists and Arrays

(added to week 8)


## Lists  - nested lists

```{r, mysize=TRUE, size='\\footnotesize'}

list(list('a','b'),list('c','d'))
```


## Arrays

Arrays are a general form a matrix, but have a higher dimension.

```{r, mysize=TRUE, size='\\tiny'}
array.1 <- array(1:8, dim=c(2,2,2)); array.1
array.1[2,2,1]
```

## Exercise: Arrays

Create an array of dimension 2 x 2 x 3, where each of the three 2 x 2 subarray (or matrix) is the Identity matrix.

## Solution: Arrays
Create an array of dimension 2 x 2 x 3, where each of the three 2 x 2 subarray (or matrix) is the Identity matrix.

```{r, mysize=TRUE, size='\\tiny'}
array(c(1,0,0,1), dim = c(2,2,3))
```


# Matrix Style Operations


## colMeans, rowSums
R contains a set of built in functions for taking the mean and sum of matrices that have been optimized for speed.

```{r colMeans}
mat1 <- matrix(1:4,ncol=2,nrow=2)
rowMeans(mat1)
colSums(mat1)
```

## Apply
For generic functions, the set of apply commands are extremely useful. They provide a mechanism for matrix style operations similar to the built in `rowMeans()` type of functions.

The apply function has three arguments: 

- matrix
- margin (rows=1, columns=2)
- function
```{r apply}
apply(mat1,2,mean)
```

## Aggregate
Another useful function is `aggregate` which can be used to compute summary statistics of dataset by a particular group. Aggregate also has three essential elements.

1. An R object
2. list of groups 
3. function

```{r agg}
aggregate(Loblolly$height,by=list(Loblolly$age),mean)
```

## Exercise: Aggregate
Earlier we wrote a function to compute the average housing price for two states, now use aggregate to compute this for all the states in the housing data set.

## Solution: Aggregate

```{r, mysize=TRUE, size='\\tiny'}
housing.prices <- read.csv(
  'http://math.montana.edu/ahoegh/teaching/stat408/datasets/HousingSales.csv',
  stringsAsFactors = F)
housing.prices.state <- aggregate(housing.prices$Closing_Price, 
                                  by=list(housing.prices$State),mean)
head(housing.prices.state)
```



# SQL

## SQLite

For this class we will use SQLite which enables users to store database files locally, but the principles are the same for querying a server-based database.

We will use a European soccer database available at [https://www.kaggle.com/hugomathien/soccer/](https://www.kaggle.com/hugomathien/soccer/) which can be downloaded with the following link: [https://www.kaggle.com/hugomathien/soccer/downloads/database.sqlite](https://www.kaggle.com/hugomathien/soccer/downloads/database.sqlite)

## Accessing Database
```{r, mysize=TRUE, size='\\footnotesize'}
library(DBI)
library(RSQLite)
## connect to a database WHICH IS STORED LOCALLY
my.database <- dbConnect(SQLite(),
  dbname="~/Google Drive/teaching/STAT408/data/database.sqlite")
dbListTables(my.database)
dbDisconnect(my.database)
```

## Identifying fields in table
```{r, mysize=TRUE, size='\\footnotesize'}
my.database <- dbConnect(SQLite(),
 dbname="~/Google Drive/teaching/STAT408/data/database.sqlite")
dbListFields(my.database, "Player")
```

## SQL commands

The most basic SQL queries have the following structure:

- **SELECT** var1name, var2name (filter columns)
- **FROM** tablename (identify table)
- **WHERE** condition1 (filter rows)
- **GROUP_BY** var3name (aggregate data)
- **HAVING** condition2 (filter aggregated data)
- **ORDER_BY** var (arrange ordering)

## SQL Query 1
Select all columns for player and view first 5 rows.
```{r, mysize=TRUE, size='\\tiny'}
kable((dbGetQuery(my.database,"SELECT * FROM Player"))[1:5,])
```

## SQL Query 2
Retain player name, weight, height for players over 200 cm
```{r, mysize=TRUE, size='\\tiny'}
kable(dbGetQuery(my.database,"SELECT player_name, height, weight FROM Player WHERE height > 200"))
```

## SQL Query 3
Compute average weight for players of 200 cm
```{r, mysize=TRUE, size='\\tiny'}
dbGetQuery(my.database,"SELECT AVG(weight) as mean_weight FROM Player WHERE height > 200")
```

## Create a database
```{r, mysize=TRUE, size='\\tiny'}
new.db <- dbConnect(RSQLite::SQLite(), ":memory:")
dbListTables(new.db)
player <- as_tibble(dbGetQuery(my.database,"SELECT * FROM Player"))
dbWriteTable(new.db, "player", player)
dbListTables(new.db)
dbDisconnect(new.db)
dbDisconnect(my.database)
```

## Additional SQL
SQL also has functionality for merging and updating tables. See the cheat sheet for more details.

## SQL Exercise
Select the average goals scored in matches in different countries from the match table

## SQL Solution
Select the average goals scored in matches in different countries from the match table

```{r, mysize=TRUE, size='\\tiny'}
my.database <- dbConnect(SQLite(),
 dbname="~/Google Drive/teaching/STAT408/data/database.sqlite")
st <- "SELECT AVG(home_team_goal + away_team_goal) as total_goals,
    country_id  FROM match GROUP by country_id"
dbGetQuery(my.database,st)
dbListTables(my.database)
```

## SQL Solution
Select the average goals scored in matches in different countries from the match table

```{r, mysize=TRUE, size='\\tiny'}
st2 <- "Create Table goals as SELECT AVG(home_team_goal + away_team_goal) as
   total_goals, country_id  FROM match GROUP by country_id "
dbGetQuery(my.database,st2)
dbGetQuery(my.database, "SELECT * from goals INNER JOIN country on goals.country_id = country.id")
dbSendQuery(my.database, "Drop Table goals")
dbDisconnect(my.database)
```