---
title: "Lab 07 - R Shiny"
output: 
  tufte::tufte_html:
    css: ../lab.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(tidyverse)
```


In this lab, you'll practice building interactive webpages using RShiny with COVID-19 case data. Rather than generating an .md file with your lab answers, your code will generate an RShiny application within an R Markdown document.

# Learning goals

-   Reviewing reshaping data
-   Creating interactive tables using RShiny
-   Creating interactive figures using RShiny


# Getting started

Each member of the team should:

- Go to the course GitHub organization (or your team page) and locate your lab repo, which should be named `lab-07-shiny-YOUR_TEAM_NAME`.
- Grab the URL of the repo, and clone it in RStudio by creating a new project from Version Control with Git.
- Open the R Markdown document `lab-07.Rmd` and Knit it.
Make sure it compiles without errors.


## Warm up

Before we introduce the data, let's warm up with some simple exercises.

-   Update the YAML of your R Markdown file with your team information and **knit** the document.
-   Commit your changes with a meaningful commit message.
-   Push your changes to GitHub.
-   Go to your repo on GitHub and confirm that your changes are visible in your Rmd **and** md files. If anything is missing, commit and push again.

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualization, and the **shiny** package to create our interactive data summaries and visualizations, and the **DT** package to render interactive tables.

```{r eval = FALSE, message = FALSE}
library(tidyverse) 
library(shiny)
library(DT)
```

## Data

We will be using a data set on COVID-19 cases. Read in the data set using the following command.

```{r, message = F, eval = FALSE, echo = TRUE}
covid_cases <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv") 
```

# Exercises


1. **Examine the data.** 

A description of the data set was not provided, other than that it is "COVID-19" data. Navigate to the Github page where the data were obtained. Use the description of this Github page and an examination of the data you loaded to write a few sentences describing this data set. What information do these data provide the user? What are the observational units and variables? What dimension is the data set? 

2. **Reshape the data.** 

Convert the data set from wide to long format.

3. **Create an Interactive Table.** 

Allow users to select the date to display in the table. Note the `DT::renderDataTable` might be useful (this is the `renderDataTable` function in the `DT` package).

4. **Create an Interactive Figure.** 

Read in a US level COVID-19 dataset and convert this to a long dataframe.
```{r, eval = FALSE, message = F}
US_data <- read_csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv") 
```
Create an interactive figure of your choice. You can consider filtering by state / province to show curves, include colors and/or faceting based on selected variables, or something else of your choosing.
