---
title: "Exam 2 Review --- Key"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

For questions 1--4, use the following data frame (called `snow`) and interpret the code. For full credit (and partial credit), write (1) what the code will return, along with (2) justification for your answer.

```{r, eval = T, echo = F}
snow <- tibble(ski_hill = c('Big Sky','Jackson Hole','Taos','Bridger Bowl','Loveland', "Vail", "Park City"),
               pass = c('Ikon','Ikon','Ikon','Powder Alliance','Powder Alliance', "Epic","Epic"),
       acres = c(5800, 2500,1294, 2000, 1800, 5317,7300),
       vertical = c(4336, 4139,3281, 2600, 2210,3450,3226 ))
snow %>% arrange(pass) %>% kable()
```


### 1.

```{r, eval=TRUE}
library(tidyverse)
snow %>% group_by(pass) %>% 
  summarize(min_size = min(acres)) %>% 
  arrange(min_size)
```

### 2.
```{r, eval=TRUE}
for (i in 1:5){
  print(snow$ski_hill[i])
    if (snow$pass[i] == 'Ikon'){
      print("is on the Ikon pass")
    } else {
      print("is not on the Ikon pass")
    }
}
```

### 3.
```{r, eval=TRUE, out.width="50%"}
snow %>% ggplot(aes(y = vertical, x = acres, shape = pass)) +
  geom_point() + ggtitle('Vertical vs. Skiable Acres')
```

### 4.
```{r, eval=TRUE}
ifelse(snow[,3] > 5000 | snow[,4] > 4000, 
       "The mountain is big or steep",
       "The mountain is not big or steep")
```



For questions 5--8, use the `snow` data frame and the `ticket_price` data frame (below). For full credit (and partial credit), write (1) what the code will return, along with (2) justification for your answer.
```{r, eval = T, echo = F}
ticket_price <- tibble(ski_hill = c('Big Sky','Bridger Bowl','Discovery'),
               pass_cost = c('1,699', '900', '525'))
ticket_price %>% kable()
```

### 5.
```{r, eval=TRUE}
typeof(ticket_price$pass_cost)
```

### 6.
```{r, eval=TRUE}
snow %>% inner_join(ticket_price, by = "ski_hill")
```


### 7.
```{r, eval=TRUE}
ticket_price %>% left_join(snow, by = "ski_hill")
```

### 8.
```{r, eval=TRUE}
library(stringr)
ticket_price %>% mutate(cost = str_replace(pass_cost, ',', ''))
```


### 9.

Finish the function. Either code or prose is acceptable.
```{r, eval=TRUE}
flip_coin <- function(num_flips){
  # Function to simulate flipping a fair coin
  # ARGS: num_flips - number of times to flip coin
  # RETURNS: vector of coin flips containing "H" or "T"
  sample(c("H", "T"), size = num_flips, replace = TRUE)
}
flip_coin(10)
```
