---
title: "Homework 3"
author: "your name here"
date: "Due **Wednesday**, February 16, 2022 at 5:00 PM in D2L"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn in both the PDF output and your R Markdown (.Rmd) file. Your .Rmd file should compile on its own if it is downloaded by your instructor.

## Load packages and data

```{r load-packages}
library(tidyverse)
library(scales)
library(fivethirtyeight)
```

## Exercises

### Exercise 1

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 2

Remove this text, and add your answer for Exercise 2 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

Remove this text, and add your answer for Exercise 3 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 4

...

### Exercise 5

...

### Exercise 6

...

### Exercise 7

...

### Exercise 8

...

### Exercise 9

...

### Exercise 10

...


## Cite Sources

Remove this text and cite the sources used on your homework here.