---
title: "Homework 5"
author: "your name here"
date: "Due Tue, Mar 8, 2022 at 5:00 PM in D2L"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dsbox) 
```

# Part I: La Quinta is Spanish for next to Denny's, Pt. 2

## Exercise 1

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.

## Exercise 2

...

## Exercise 3

...

## Exercise 4

...

## Exercise 5

...


## Exercise 6

```{r}
haversine <- function(long1, lat1, long2, lat2, round = 3) {
  # convert to radians
  long1 = long1 * pi / 180
  lat1  = lat1  * pi / 180
  long2 = long2 * pi / 180
  lat2  = lat2  * pi / 180
  
  R = 6371 # Earth mean radius in km
  
  a = sin((lat2 - lat1)/2)^2 + cos(lat1) * cos(lat2) * sin((long2 - long1)/2)^2
  d = R * 2 * asin(sqrt(a))
  
  return( round(d,round) ) # distance in km
}
```

## Exercise 7


...

Add exercise headings as needed.

# Part II: History of Baseball

```{r, eval = FALSE, message = F}
player <- read_csv("https://math.montana.edu/shancock/data/player.csv")
all_star <- read_csv("https://math.montana.edu/shancock/data/all_star.csv")
salary <- read_csv("https://math.montana.edu/shancock/data/salary.csv")
```

## Exercise 11

...

Add exercise headings as needed.


# Cite Sources

Remove this text and cite the sources used on your homework here.