---
title: "Homework 8"
author: "your name here"
date: "Due Wed, Apr 13, 2022 at 11:00 PM in D2L"
urlcolor: blue
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(openintro)
```


# Exercises

## Baby weight vs. smoking

The following problems use the `ncbirths` data set explored in Lab 08. For these exercises, we will use the entire data set (not just white mothers as in the lab).

Consider the possible relationship between a mother's smoking habit and the weight of her baby.

1.  Create a version of the `ncbirths` dataset omitting observations where there are NAs for `habit`.
    You can call this version `ncbirths_habitgiven`.
    
2.  Plotting the data is a useful first step because it helps us quickly visualize trends, identify strong associations, and develop research questions. Create an appropriate plot displaying the relationship between `habit` and `weight`.
    What does the plot highlight about the relationship between these two variables?

3.  Calculate the mean `weight` in each of the `habit` groups. How do these two means compare?

There is an observed difference in sample means, but is this difference statistically significant? That is, if there really was no relationship between a mother's smoking habit and the weight of her baby, is this difference in sample means unlikely to occur just by chance?
In order to answer this question you will conduct a hypothesis test.

4.  Using standard statistical notation, write the hypotheses for testing if the average weights of babies born to smoking and non-smoking mothers are different. Hint: Mathematical notation can be created in your .Rmd file using LaTeX code, which is surrounded by `$`. For instance, the alternative hypothesis displayed before Exercise 3 in Lab 08 was created with the code
`$H_A: \mu \neq 7.43$`.

5.  Use bootstrapping to construct a 95% confidence interval for the difference between the average weights of babies born to smoking and non-smoking mothers.

6.  Use your interval from the previous exercise to state a conclusion to the hypothesis test.

## Extra Credit

Write and run code that will carry out a randomization test for these data. If you are unfamiliar with a randomization test for this scenario, or you just want a refresher, review [Section 6.3.1](https://mtstateintrostats.github.io/IntroStatTextbook/inference-num.html#rand2mean) of the STAT 216 online textbook. 

You may write code using only base R functions, or you can explore the `tidymodels` or `infer` packages, which include functions to run simulation-based hypothesis tests.
