---
title: "Homework 6"
author: "your name here"
date: "Due Thur, Mar 24, 2022 at 11:00 PM in D2L"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
```

The data we will use for this homework contain single bike trips from March of 2017
for the Capital BikeShare system in Washington, D.C.

```{r load-data, cache = TRUE, message = FALSE}
bikes <- read_csv("https://math.montana.edu/shancock/data/biketrips2017.csv")
```

_Notice_: This is a large file, so it takes RStudio a few minutes to import the data.
In order to avoid waiting for this data to load each time you knit the document,
I have added the [cache option](https://bookdown.org/yihui/rmarkdown-cookbook/cache.html)
to the R chunk above that reads in the data.


# Exercises

## 1.
Use the `str()` function to summarize the data set. 
What does each column represent? What about each row? 

## 2.
Describe the difference between `substr()` and `strsplit()`. 

## 3.
Use one of the functions described in Exercise 2
to create a new variable for the hour a bike trip began.
Then use the `count()` function to compute the number of trips starting
at each hour.

## 4.
Now, instead of using strings, we'll use the `lubridate` package to
extract parts of dates and times. Examine the `lubridate` cheatsheet
[here](https://rawgit.com/rstudio/cheatsheets/main/lubridate.pdf).
What function from this package could you use to extract the day of the month?

## 5.
Use the function you chose in Exercise 4 to create a new variable called `day`
that extracts the day of the month.
Then use this variable to compute how many trips were made for each of the
31 days in the month of March.
(The cases that started and ended on different days have been already
filtered out of the data set.)

## 6.
Create a new variable that contains the trip time.

## 7.
What percentage of bike rentals last more than an hour?

## 8.
Create a figure to plot the trip time as a function of day of week
(Sunday, Monday, etc...).
Make sure to include an informative title and
sort the weekday into chronological ordering.
Write a few sentences describing the key features of the plot.

