---
title: "Lab 08 - Smoking during pregnacy"
subtitle: "Simulation based inference and bootstrapping"
output: 
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: ../lab.css
link-citations: yes
---

```{r include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

In 2004, the state of North Carolina released a large data set containing information on births recorded in this state.
This data set is useful to researchers studying the relation between habits and practices of expectant mothers and the birth of their children.
We will work with a random sample of observations from this data set.

# Learning goals

-   Using bootstrapping methods to estimate sampling variability
-   Calculating confidence intervals using percentiles of a bootstrap distribution
-   Conducting hypothesis tests via simulation
-   Interpreting confidence intervals and results of hypothesis tests in context of the data

# Getting started

Each member of the team should:

- Go to the course GitHub organization (or your team page) and locate your lab repo, which should be named `lab-08-inference-smoking-YOUR_TEAM_NAME`.
- Grab the URL of the repo, and clone it in RStudio by creating a new project from Version Control with Git.
- Open the R Markdown document `lab-08.Rmd` and Knit it.
Make sure it compiles without errors.
The output will be in the file markdown `.md` file with the same name.


## Warm up

Let's warm up with some simple exercises.

-   Update the YAML of your R Markdown file with your team information and **knit** the document.
-   Commit your changes with a meaningful commit message.
-   Push your changes to GitHub.
-   Go to your repo on GitHub and confirm that your changes are visible in your Rmd **and** md files. If anything is missing, commit and push again.

## Packages

We'll use the **tidyverse** package for much of the data wrangling and visualization, and the data lives in the **openintro** package.
You can load them by running the following in your Console:

```{r}
library(tidyverse) 
library(openintro)
```

## Data

The data can be found in the **openintro** package, and it's called `ncbirths`.
Since the dataset is distributed with the package, we don't need to load it separately; it becomes available to us when we load the package.
You can find out more about the dataset by inspecting its documentation, which you can access by running `?ncbirths` in the Console or using the Help menu in RStudio to search for `ncbirths`.
You can also find this information [here](https://www.openintro.org/data/index.php?data=ncbirths).

# Set a seed!

In this lab we'll be generating random samples.
The last thing you want is those samples to change every time you knit your document.
So, you should set a seed.
There's an R chunk in your R Markdown file set aside for this.
Locate it and add a seed.
Make sure all members in a team are using the same seed so that you don't get merge conflicts and your results match up for the narratives.

# Exercises

1.  What are the cases in this data set? How many cases are there in our sample?

The first step in the analysis of a new dataset is getting acquainted with the data.
Make summaries of the variables in your dataset, determine which variables are categorical and which are numerical.
For numerical variables, are there outliers?
If you aren't sure or want to take a closer look at the data, make a graph.

## Baby weights

A 1995 study suggests that average weight of Caucasian babies born in the US is 3,369 grams (7.43 pounds).[^lab-08-inference-smoking-1]
In this dataset we only have information on mother's race, so we will make the simplifying assumption that babies of Caucasian mothers are also Caucasian, i.e. `whitemom = "white"`.

We want to evaluate whether the average weight of Caucasian babies has changed since 1995.

Our null hypothesis should state "there is nothing going on", i.e., no change since 1995: $H_0: \mu = 7.43$ pounds.

Our alternative hypothesis should reflect the research question, i.e., some change since 1995.
Since the research question doesn't state a direction for the change, we use a two sided alternative hypothesis: $H_A: \mu \ne 7.43$ pounds.

3.  Create a filtered data frame called `ncbirths_white` that contain data only from white mothers.
    Then, calculate the mean of the weights of their babies.

4.  Are the conditions necessary for conducting simulation-based inference satisfied?
    Explain your reasoning.

We will use simulation-based methods to calculate a confidence interval for
the true mean weight of Caucasian babies in North Carolina in 2004, as well as test the hypotheses stated above.

## Confidence interval

5.  Using the R script file we created in class as a guide, write and run code to generate a bootstrap distribution of 10,000 sample mean weights from your `ncbirths_white` data set. Then:

- produce a (well-labeled!) histogram of this distribution, 
- use the distribution to generate a 95% confidence interval for $\mu$, and
- write a sentence interpreting your confidence interval in context of the problem.

_The above exercise will take a while --- make sure every team member is contributing to the solution. That may mean getting up and all looking at the same computer!_


## Hypothesis test

Let's first discuss how a hypothesis test using simulation would work.
Our goal is to simulate a null distribution of sample means that is centered at the null value of 7.43 pounds.
In order to do so, we

-   take a bootstrap sample from the original sample,
-   calculate this bootstrap sample's mean,
-   repeat these two steps a large number of times to create a bootstrap distribution of means centered at the observed sample mean,
-   shift this distribution to be centered at the null value by subtracting / adding X to all bootstrap mean (X = difference between mean of bootstrap distribution and null value), and
-   calculate the p-value as the proportion of bootstrap samples that yielded a sample mean at least as extreme as the observed sample mean.

Note, we already did the first three steps in the previous section!

7.  Use the steps above to run the appropriate hypothesis test, visualize the null distribution, calculate the p-value, and interpret the results in context of the data and the hypothesis test. 

_Every team member should be contributing to the solution for the above exercise!_



# Attribution

This lab is adapted from material in the [Data Science in a Box](https://datasciencebox.org/index.html) course by Mine Çetinkaya-Rundel licensed under a [Creative Commons Attribution Share Alike 4.0 International](https://creativecommons.org/licenses/by-sa/4.0/).
Visit [here](https://github.com/rstudio-education/datascience-box/blob/master/LICENSE.md) for more information about the license.



[^lab-08-inference-smoking-1]: Wen, Shi Wu, Michael S. Kramer, and Robert H. Usher.
    "Comparison of birth weight distributions between Chinese and Caucasian infants." American Journal of Epidemiology 141.12 (1995): 1177-1187.
