---
title: "HW 04 - Style, Debugging and Functions"
output: 
  tufte::tufte_html:
    css: ../hw.css
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

```{r include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  out.width = "80%",
  fig.asp = 0.618,
  fig.width = 10,
  dpi = 300
)
```

# Getting started

* Open RStudio. Navigate to File > New Project..., and select "New Directory", then "New Project". Name your directory something like "hw-04" and save it as a subdirectory of your STAT 408 folder on your computer. Click "Create Project".
* Download the [R Markdown document `hw-04.Rmd`](hw-04.Rmd) from the course calendar and save it to your newly created "hw-04" folder.
* Open the `hw-04.Rmd` document in RStudio. Update the YAML, changing the author name to your name, and **knit** the document to PDF.
* Make sure it compiles without errors. The output will be in the `.pdf` file with the same name in the same directory.

# RStudio tips and tricks

1. Go to the RStudio Tips twitter account, [https://twitter.com/rstudiotips](https://twitter.com/rstudiotips),
and find one tip that looks interesting and helpful for our class.
Practice using it! Then post the tip and a couple sentences
describing why you like it or how you used it in our D2L Discussion 
"R/RStudio Tips & Tricks" topic.


# RStudio addins

```{marginfigure}
Other options for automated styling in RStudio include the built-in
Code > Reformat Code option, or the 
[`formatR`](https://yihui.org/formatr/) package.
```

We spent the better part of a class period describing the 
features of R style guidelines,
but anything with rules (like style guidelines) can easily be automated.
And since RStudio is created by programmers, there are tools within RStudio to
help us adhere to a particular style. One option for automated styling is
the [styler](https://styler.r-lib.org/index.html) addin.

Read the following two pages about RStudio addins and the styler add-in:

* [Overview of RStudio add-ins](https://towardsdatascience.com/rstudio-addins-or-how-to-make-your-coding-life-easier-6b627c7b2240)
* [Styler](https://www.tidyverse.org/blog/2017/12/styler-1.0.0/)

Install the styler addin, and then complete the following exercises.

Consider the following code:

```{r, eval=FALSE}
data(iris) #read in the data set
r = cor(iris$Sepal.Length,iris$Sepal.Width);r
iris%>%group_by(Species) %>% summarize(r.sepal=cor(Sepal.Length,Sepal.Width), r.petal=cor(Petal.Length,Petal.Width),mean.sepal=mean(Sepal.Length),mean.petal=mean(Petal.Length)) #correlation by species
iris%>%ggplot(aes(x=Sepal.Length,y=Sepal.Width,col=Species))+
  geom_point()+theme_minimal()+ggtitle(`Sepal Width vs Height by Species`)
#here is some basic math
1*2^5
#and a function
triangle_area=function ( b,h ){#calculate area of a triangle with base b and height h
  (1/2)*b*h}
```

2. Copy and paste the code above into an R chunk in your `hw-04.Rmd` file. 
Select the code and run styler addin on the selection. 

3. Create a bulleted list describing which features in the code changed
and how they were styled, e.g.,

* changed `=` to `<-` when used as assignment
* ...

4. Are there any tidyverse style guidelines that the styler addin missed?
If so, what?

# Debugging Part I

5. The goal of this code is to create a figure of age by passenger class among passengers that were on the Titanic. There are a few bugs in the code; identify and fix them. List all the things that you changed.
```{r, eval=FALSE}
titanic == read_csv("http://math.montana.edu/ahoegh/teaching/stat408/datasets/titanic.csv")

titanic 
  %>% filter(!is.na(Age)) %>% # removed passengers without age
  mutate(Pclass = factor(Pclass)) %>% # changed class to factor
  ggplot(y = Age, x = Pclass)) %>%
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(color = Sex) +
  theme_bw() + 
  xlab(Passenger Class) +
  ggtitle('Passenger age by class and gender on Titanic')
```


# Writing functions

There is a classic probability problem called "the hat problem": 

> Suppose $n$ people go to a fancy restaurant.
Each person is wearing a hat and checks his/her
hat at the door as he/she arrives. The hat-check
attendant gets tipsy throughout the evening,
and returns a random hat to each person as
they leave. The patrons leave in a random order.
What is the probability that no one gets his or her own hat back? 

The goal of this part of the homework is to write
a function that will take an argument $n$ and return
the estimated probability that no one gets their own hat back.

In order to estimate this probability, we need to simulate
the random process many times, and then calculate the
proportion of times no one got their hat back.

The code below will simulate one trial of the hat process
for $n = 20$ and return the number of people who got their
hat back.

```{r}
hats <- sample(1:20, 20, replace = FALSE)
heads <- sample(1:20, 20, replace = FALSE)
sum(hats == heads)
```

Try running the code and make sure you understand each line
of the code before proceeding.

6. Fill in the for loop below to simulate the hat process
1000 times for $n = 20$, storing the number of matches
from each iteration in the object `n_matches`.
```{r}
n_matches <- vector("integer", 1000)
for(i in seq_along(n_matches)) {
  # add code here to simulate hat process
  # store the result from each iteration
  # in n_matches
}
```

```{marginfigure}
**Hint:** The code `n_matches == 0` will generate a
logical vector. Since logical vectors are treated
as 0's (FALSE) and 1's (TRUE), `mean(n_matches == 0)`
will return the proportion of TRUEs in the vector.
```

7. Use the `n_matches` object created in the last exercise
to estimate the probability of zero hat matches when
$n = 20$.

8. Now, create a function called `FindHatProbability` that
takes arguments `n` (number of hats) and `reps` (number
of times to simulate the process), and returns an estimated
probability of zero hat matches. Set the default value
for `reps` to 10000.

```{marginfigure}
It can be shown that the limit of this probability
as $n$ goes to infinity is $1/e$!
```


9. Use your function for the previous exercise to estimate
the probability of zero matches for $n = 5, 10, 20, 50$.
How does this probability change as $n$ increases?

# Debugging Part II

```{marginfigure}
For a short video of the Monty Hall problem see [from 21 with Kevin Spacey](https://www.youtube.com/watch?v=Zr_xWfThjJ0) or 
[from numb3rs tv show](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=0ahUKEwjLwcvG1rTWAhUrsVQKHWq9CbEQyCkIKjAA&url=https%3A%2F%2Fwww.youtube.com%2Fwatch%3Fv%3DP9WFKmLK0dc&usg=AFQjCNEYUi91dn4_xNLXQr9pESAdn9aApQ).
```

Another fun probability problem is the "Monty Hall problem".
Deriving the probabilities in the hat problem or the
Monty Hall problem requires some advanced probability
knowledge, but estimating the probabilities through
simulation takes only a bit of coding!


10. Debug the following function, by rewriting the function below
and demonstrating that the function calls specified below return the 
correct answer. 
```{r, error= TRUE}
MontyHallMonteCarlo <- function(num.sims, print){
  # Function to simulate Monty Hall winning probability when switching doors
  # ARGS: number of simulations (as integer or double), print command
  #       that accepts TRUE or FALSE as to whether to print simulation results
  # Returns: list containing winning probability and (if print = TRUE)
  #          vector of results with strings "Win" or "Lose" for each simulation
  if (!num.spins %% 1 == 0) stop('Please enter an integer or double')
  results <- rep(FALSE,num.sims)
  for (i in 1:num.sims){
    # randomly choose door with car
    car.door <- sample(3,1)
    # randomly choose door for participant to select
    select.door <- sample(1,3)
    # you win when switching if the door with a car is not the
    # one you initally selected
    if (car.door = select.door) {
      results <- FALSE
    }
  }
  win.prob <- mean(results)
  ifelse(print, return(list(win.prob,results)),return(list(win.prob))
}

MonteHallMonteCarlo(8.1,print=T)
MonteHallMonteCarlo('8.1',print=T)
MonteHallMonteCarlo(8,print=T)
MonteHallMonteCarlo(10000,print=F)
```


# Cite sources

Write the sources you used to complete this assignment at the end of your .Rmd document, adhering to the "Guidance on Citing Sources" bullet points in the [collaboration policy section on our course syllabus](https://stat408-s22.github.io/#Policy_on_collaboration_and_academic_misconduct).

