---
title: "Homework 4"
author: "your name here"
date: "Due Wed, Mar 2, 2022 at 5:00 PM in D2L"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Instructions

Use this .Rmd file as a template for your homework. Please use D2L to turn 
in both the Knitted PDF output and your R Markdown file. Your .Rmd file 
should compile on its own if it is downloaded by your instructor.

## Load packages

```{r load-packages, message = FALSE}
library(tidyverse)
```

# RStudio tips and tricks

## Exercise 1

Remove this text, and copy and paste your D2L discussion post here.


# RStudio add-ins.

## Exercise 2

```{r, eval = FALSE}
# copy and paste code to style here
```

Remove this text, and add your answer for Exercise 2 here.


## Exercise 3

Remove this text, and add your answer for Exercise 3 here.


## Exercise 4

Remove this text, and add your answer for Exercise 4 here.


# Debugging Part I

## Exercise 5

The goal of this code is to create a figure of age by passenger 
class among passengers that were on the Titanic. There are a
few bugs in the code; identify and fix them.
List all the things that you changed.
```{r, eval=FALSE}
titanic == read_csv("http://math.montana.edu/ahoegh/teaching/stat408/datasets/titanic.csv")

titanic 
  %>% filter(!is.na(Age)) %>% # removed passengers without age
  mutate(Pclass = factor(Pclass)) %>% # changed class to factor
  ggplot(y = Age, x = Pclass)) %>%
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(color = Sex) +
  theme_bw() + 
  xlab(Passenger Class) +
  ggtitle('Passenger age by class and gender on Titanic')
```

After you have debugged the code above, change the option `eval = FALSE`
to `eval = TRUE` in the R chunk to demonstrate that the code runs.


# Writing functions

## Exercise 6

Remove this text, and write your answer for Exercise 6 here.
Add R chunks as needed.

## Exercise 7

...

## Exercise 8

...

## Exercise 9

...


# Debugging Part II

## Exercise 10

Debug the following function, by rewriting the function below
and demonstrating that the function calls specified below
return the correct answer.

```{r, eval=FALSE}
MontyHallMonteCarlo <- function(num.sims, print){
  # Function to simulate Monty Hall winning probability when switching doors
  # ARGS: number of simulations (as integer or double), print command
  #       that accepts TRUE or FALSE as to whether to print simulation results
  # Returns: list containing winning probability and (if print = TRUE)
  #          vector of results with strings "Win" or "Lose" for each simulation
  if (!num.spins %% 1 == 0) stop('Please enter an integer or double')
  results <- rep(FALSE,num.sims)
  for (i in 1:num.sims){
    # randomly choose door with car
    car.door <- sample(3,1)
    # randomly choose door for participant to select
    select.door <- sample(1,3)
    # you win when switching if the door with a car is not the
    # one you initally selected
    if (car.door = select.door) {
      results <- FALSE
    }
  }
  win.prob <- mean(results)
  ifelse(print, return(list(win.prob,results)),return(list(win.prob))
}

MonteHallMonteCarlo(8.1,print=T)
MonteHallMonteCarlo('8.1',print=T)
MonteHallMonteCarlo(8,print=T)
MonteHallMonteCarlo(10000,print=F)
```

After you have debugged the code above, change the option `eval = FALSE`
to `eval = TRUE` in the R chunk to demonstrate that the code runs.


# Cite Sources

Remove this text and cite the sources used on your homework here.